Covariate analysis of HBCC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (DLPFC with CQN normlaisation)

### 2017-06-30 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Minor modifications

knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
                                 parentId ="syn7501880",
                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn7501880';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (ACC and DLPFC)';

@@

### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=15}
# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueSCZ_ACC",
                                     "Dx.TissueControl_DLPFC-Dx.TissueControl_ACC",
                                     "0.5*Dx.TissueSCZ_DLPFC+0.5*Dx.TissueSCZ_ACC-0.5*Dx.TissueControl_DLPFC-0.5*Dx.TissueControl_ACC",
                                     "0.5*Dx.TissueSCZ_DLPFC+0.5*Dx.TissueControl_DLPFC-0.5*Dx.TissueSCZ_ACC-0.5*Dx.TissueControl_ACC"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR) 
names(DE) = gsub('Dx.Tissue', '', colnames(contrast))
names(DE)[9:10] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC')

@@

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, ref.Dx != 'Other', to.Dx != 'Other') %>%
  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
  group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), ref.Dx != 'Other', to.Dx != 'Other') %>%
  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
  group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))

@@

### Gene set enrichment analysis
Using Fisher's exact test for gene set enrichment analysis
```{r enrich.de, fig.height=16, fig.width=16}
diffexp.genes = DE %>%
  dlply(.(ref.Dx, ref.Tissue, to.Dx, to.Tissue), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
backgroundGenes = unique(DE$hgnc_symbol)

>>>>

knit2synapse::knitToFolderEntity(file = "./CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
                                 parentId = "syn9872607",
                                 entityName = 'ACC and DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn9872607';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (ACC and DLPFC)';

@@

### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=15}
# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueSCZ_ACC",
                                     "Dx.TissueControl_DLPFC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_DLPFC+Dx.TissueSCZ_ACC-Dx.TissueControl_DLPFC-Dx.TissueControl_ACC",         
                                     "Dx.TissueSCZ_DLPFC+Dx.TissueControl_DLPFC-Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC-Dx.TissueSCZ_ACC+Dx.TissueControl_ACC"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR) 
names(DE) = gsub('Dx.Tissue', '', colnames(contrast))
names(DE)[5:7] = c('SCZ_ALL-Control_ALL', 'ALL_DLPFC-ALL_ACC', 'ALL_ALL-ALL_ALL')

@@

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, ref.Dx != 'Other', to.Dx != 'Other') %>%
  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue, logFC) %>%
  group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), ref.Dx != 'Other', to.Dx != 'Other') %>%
  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue, logFC) %>%
  group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))

@@

```{r diffExp1, fig.height=10, fig.width=15}
postAdjustCovars = c('Tissue', 'Dx', 'Institution', 'Gender', 'RIN', 'IntronicRate')

# Post adjusted design matrix
COV1 = COVARIATES
COV1 = tidyr::separate(COV1, Dx.Tissue, c('Dx', 'Tissue'), sep = '_') %>%
  dplyr::mutate(Dx = factor(Dx, levels = c('SCZ', 'Control', 'Other')), Tissue = factor(Tissue))

DESIGN = getDesignMatrix(COV1[,postAdjustCovars,drop=F],Intercept = F)$design
DESIGN = DESIGN[,linColumnFinder(DESIGN)$indepCols]

# Re-calculate voom weights with correlation of random effects
VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero,   
                            design = DESIGN, plot=F,
                            block = COV1$Individual_ID, 
                            correlation = correlation$cor)
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E

# Fit contrast
contrast = makeContrasts(contrasts=c("TissueACC-TissueDLPFC",
                                     "DxSCZ-DxControl"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE1 = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR) 
names(DE1) = c('ACC-DLPFC', 'SCZ-CONTROL')

DE1 = DE1 %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Direction = sign(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(Ensemble2HGNC) 
DE1$Direction[DE1$adj.P.Val > 0.05 | abs(DE1$logFC) < log2(1.2)] = 'NONE'

tmp = DE1 %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
  group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))

tmp1 = DE1 %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
  group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))

p = DE1 %>%
  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(Comparison~., scales = 'free_y')
p = p + ggtitle('Differential Expression between Schizophrenia and Control')
p
```

### Gene set enrichment analysis
Using Fisher's exact test for gene set enrichment analysis
```{r enrich.de, fig.height=16, fig.width=16}
diffexp.genes.up = DE %>%
  dlply(.(ref.Dx, ref.Tissue, to.Dx, to.Tissue), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC > 0] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
names(diffexp.genes.up) = paste(names(diffexp.genes.up), 'UP', sep = '.')

diffexp.genes.down = DE %>%
  dlply(.(ref.Dx, ref.Tissue, to.Dx, to.Tissue), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC < 0] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
names(diffexp.genes.down) = paste(names(diffexp.genes.down), 'DOWN', sep = '.')

diffexp.genes.up$ACC.DLPFC.UP = dplyr::filter(DE1, adj.P.Val <= 0.05, logFC > 0, Comparison == 'ACC-DLPFC') %>%
  dplyr::select(hgnc_symbol) %>%
  unique() %>% unlist() %>% as.character()
diffexp.genes.down$ACC.DLPFC.DOWN = dplyr::filter(DE1, adj.P.Val <= 0.05, logFC < 0, Comparison == 'ACC-DLPFC') %>%
  dplyr::select(hgnc_symbol) %>%
  unique() %>% unlist() %>% as.character()
diffexp.genes.up$ACC.DLPFC.UNCHANGED = dplyr::filter(DE1, adj.P.Val > 0.05, Comparison == 'ACC-DLPFC') %>%
  dplyr::select(hgnc_symbol) %>%
  unique() %>% unlist() %>% as.character()

diffexp.genes = c(diffexp.genes.up, diffexp.genes.down)
diffexp.genes = diffexp.genes[c("Control.ACC.SCZ.ACC.UP", "Control.ACC.SCZ.ACC.DOWN",
                                "Control.DLPFC.SCZ.DLPFC.UP", "Control.DLPFC.SCZ.DLPFC.DOWN",
                                "ACC.DLPFC.UP", "ACC.DLPFC.DOWN", "ACC.DLPFC.UNCHANGED")]
diffexp.genes$Control.SCZ.ACC.DLPFC.UP = setdiff(diffexp.genes$Control.ACC.SCZ.ACC.UP, diffexp.genes$Control.DLPFC.SCZ.DLPFC.UP)
diffexp.genes$Control.SCZ.ACC.DLPFC.DOWN = setdiff(diffexp.genes$Control.ACC.SCZ.ACC.DOWN, diffexp.genes$Control.DLPFC.SCZ.DLPFC.DOWN)
diffexp.genes$Control.SCZ.DLPFC.ACC.UP = setdiff(diffexp.genes$Control.DLPFC.SCZ.DLPFC.UP, diffexp.genes$Control.ACC.SCZ.ACC.UP)
diffexp.genes$Control.SCZ.DLPFC.ACC.DOWN = setdiff(diffexp.genes$Control.DLPFC.SCZ.DLPFC.DOWN, diffexp.genes$Control.ACC.SCZ.ACC.DOWN)
diffexp.genes$Control.SCZ.DLPFC.ACC.BOTH.UP = intersect(diffexp.genes$Control.DLPFC.SCZ.DLPFC.UP, diffexp.genes$Control.ACC.SCZ.ACC.UP)
diffexp.genes$Control.SCZ.DLPFC.ACC.BOTH.DOWN = intersect(diffexp.genes$Control.DLPFC.SCZ.DLPFC.DOWN, diffexp.genes$Control.ACC.SCZ.ACC.DOWN)
diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
backgroundGenes = unique(DE$hgnc_symbol)

### 2017-06-30 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Minor modifications

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=8}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
tmp = COVARIATES.CORRELATION$ESTIMATE
tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0
h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate')
draw(h, heatmap_legend_side = 'left', padding = unit(c(10,2,2,10), "mm"))
```

@@

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue1) %>%
  group_by(Comparison, Tissue1) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue1) %>%
  group_by(Comparison, Tissue1) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))

@@

### Gene set enrichment analysis
Using Fisher's exact test for gene set enrichment analysis
```{r enrich.de, fig.height=16, fig.width=16}
diffexp.genes = DE %>%
  dlply(.(Comparison, Tissue), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
backgroundGenes = unique(DE$hgnc_symbol)

@@

# Perform enrichment analysis
enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
    }, x, backgroundGenes, .parallel = TRUE) %>%
    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
}, genesets.to.test, backgroundGenes) %>%
  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
  tidyr::separate(Comparison, c('Name', 'Region'), sep = '\\.') %>%
  dplyr::rename(Comparison = Name, Category = .id)
```

>>>>

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=8}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
tmp = COVARIATES.CORRELATION$ESTIMATE
tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0
fdr = COVARIATES.CORRELATION$PVAL
fdr = format(fdr, digits=1)
h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate',
            cell_fun = function(j,i,x,y,wifht,height,fill){
                grid.text(fdr[i,j], x, y)
              })
draw(h, heatmap_legend_side = 'left', padding = unit(c(10,2,2,10), "mm"))
```

@@

tmp = de.hbcc %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue1, logFC) %>%
  group_by(Comparison, Tissue1) %>%
  dplyr::summarise(FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])),
                   FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])))

tmp1 = de.hbcc %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue1, logFC) %>%
  group_by(Comparison, Tissue1) %>%
  dplyr::summarise(FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > log2(1.2)])),
                   FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < -log2(1.2)])))

@@

### Gene set enrichment analysis
Using Fisher's exact test for gene set enrichment analysis
```{r enrich.de, fig.height=16, fig.width=16}
diffexp.genes = de.hbcc %>%
  dplyr::mutate(Direction = sign(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP'))) %>%
  dlply(.(Comparison, Tissue1, Direction), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
backgroundGenes = unique(DE$hgnc_symbol)

@@

# Perform enrichment analysis
enrichment.results = plyr::llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
    }, x, backgroundGenes, .parallel = TRUE) %>%
    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
}, genesets.to.test, backgroundGenes) %>%
  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
  tidyr::separate(Comparison, c('Name', 'Region'), sep = '\\.') %>%
  dplyr::rename(Comparison = Name, Category = .id)
```
